High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-23 doc/asplos2013/results.tex
Some minor edits to the results

\section{Application Evaluation}
\label{sec:apps}
Legion is a high-level runtime system that has been implemented on top
of our interface\cite{Legion12}.  
%The Legion runtime is distributed so 
%that there is an independent scheduler for each processor in the machine.  
The Legion programming model is built around the abstraction of 
{\em logical regions} which express locality and independence properties 
of data.  Computation in Legion is organized into a tree of tasks where 
each task must specify which logical regions will be accessed.
When executing a Legion task, the Legion runtime receives requests
to execute sub-tasks along with their region requirements.
This stream of tasks with region requirements is analogous to a stream
of instructions with register requirements that are executed by 
a hardware processor.  Hardware out-of-order processors are designed to
run ahead of the actual execution of a stream of instructions
to ensure that the processor's pipeline is fully utilized.  Similarly, in a
Legion application, for every processor there is an instance of the Legion runtime
that takes a stream of tasks with region requirements and asynchronously
runs ahead of the actual execution using a {\em software-out-of-order} (SOOP)
scheduler.  The SOOP leverages our interface by asynchronously issuing task launches, copy
operations, and synchronization operations with the dependencies between them expressed
as events.  The full details of the SOOP are beyond the scope of this paper
and are described in \cite{Legion12}.

@@

{\em Circuit} is simulation of an electric circuit that is described as an
irregular graph of integrated
circuit components connected by wires.  The graph is partitioned
and distributed across the machine.  The computation for each time step consists
of three phases that calculate currents on the wires, distribute charge between
nodes, and update the voltages of all ndoes.  In the distribute charge
phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
are used to allow each partition to perform reductions in parallel and locally in GPU
framebuffer memeory before merging all the reductions to a global instance
residing in GASNet memory.  Without reduction instances this phase
couldn't be performed using GPUs.

@@

We instrumented the heterogenous implementation of our interface to 
capture event usage information.  Figure~\ref{fig:eventlife} shows
a timeline of the execution of the Fluid application on 16 nodes.
The dynamic events line measures the total number of event creations.
A large number of events are created - over 260,000 in less than 15
seconds - and allocating separate storage for every event would clearly
be difficult for long-running applications.  The {\em live events} line
measures the number of events that are live at any given time - after
the last operation on an event, a reference counting implementation would
be able to recover the storage associated with the event.  Such an
algorithm would reduce the storage needed by over 10X at the cost of
computation and communication overhead for performing the reference 
counting.  As discussed in Section~\ref{subsec:eventimpl}, our implementation
requires storage that grows with the number of {\em untriggered events}, a number
that is 10X smaller than even the maximal live event count.  The actual storage
requirements of our implementation are shown by the {\em physical events} line,
which is larger than the peak number of untriggered events because of the
decision to always create an event owned by the node from which the request
was made.  Despite this inefficiency, our implementation uses 5X less storage
than a reference counting implementation and avoids any overhead related to 
reference counting.  These savings would likely be even more dramatic for longer 
runs of the application, as the number of live events is steadily growing as the
application runs, while the peak number of generational events needed appears to 
occur during the startup of the application.

@@

We also instrumented our heterogeneous implementation to profile the usage of 
locks in real applications.  We measured how many locks were used in the application.  For
each lock we measured the number of lock requests and the number of times that the lock
had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results 
for the Circuit and AMR application, both running on 16 nodes.  The Circuit application uses
a larger number of locks (3336 vs. 1393), but most of them (86\%) are only used on the node on 
which they were created.  The AMR application exhibits more varied locking behavior, and in 
particular has locks that are requested more often than they are transferred.  Without the ability
for locks to migrate, more of those requests would have required inter-node communication.

@@

Our last set of experiments demonstrates that the asynchronous nature of our interface
is essential to the high performance of the Legion runtime system.  The Legion
implementations of our three applications have already been shown to outperform 
pre-existing implementations\cite{Legion12}.  To quantify the benefit of the asynchronous
nature of our interface, we implemented versions of our applications in
Legion in a bulk-synchronous style to compare against our existing implementations.  
These bulk-synchronous versions of still run on top of 
the Legion runtime, but invoke blocking operations between computational phases,
communication phases, and synchronization.

@@

These results show that the asynchronous nature of our interface allows
the Legion runtime system to run ahead of actual execution which
provides significant latency hiding capability.  
The benefit of this latency hiding grows as the number of
nodes increases and as the computation-to-communication ratio decreases.
In all cases, the overhead of the bulk-synchronous
implementations grew with the node count.  As we continue to scale
applications to larger and larger machines, the ability of runtimes like Legion to run ahead
will be essential to hiding the large latencies inherent in such machine architectures.
The fully asynchronous nature of the interface presented in this paper
will be crucial for supporting this ability.

>>>>

\section{Application Evaluation}
\label{sec:apps}
Legion is a high-level runtime system that has been implemented on top
of our interface\cite{Legion12}.  
%The Legion runtime is distributed so 
%that there is an independent scheduler for each processor in the machine.  
The Legion programming model is built around the abstraction of 
{\em logical regions} which express locality and independence properties 
of data.  Computation in Legion is organized into a tree of tasks where 
each task must specify which logical regions will be accessed.
When executing a Legion task, the Legion runtime receives requests
to execute sub-tasks along with their region requirements.
This stream of tasks with region requirements is analogous to a stream
of instructions with register requirements that are executed by 
a hardware processor.  Hardware out-of-order processors are designed to
run ahead of the actual execution of a stream of instructions
to ensure that the processor's pipeline is fully utilized.  Similarly, in a
Legion application, for every processor there is an instance of the Legion runtime
that takes a stream of tasks with region requirements and asynchronously
runs ahead of the actual execution using a scheduler called a {\em Software Out-of-Order Processor} 
(SOOP).  The SOOP leverages our interface by asynchronously issuing task launches, copy
operations, and synchronization operations with the dependencies between them expressed
as events.  The full details of the SOOP are beyond the scope of this paper
and are described in \cite{Legion12}.

@@

{\em Circuit} is a simulation of an integrated circuit that is described as an
irregular graph of components connected by wires.  The graph is partitioned
and distributed across the machine.  The computation for each time step consists
of three phases that calculate currents on the wires, distribute charge between
nodes, and update the voltages of all nodes.  In the distribute charge
phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
are used to allow each partition to perform reductions in parallel and locally in GPU
framebuffer memeory before merging all the reductions to a global instance
residing in GASNet memory.  Without reduction instances this phase
couldn't be performed using GPUs.

@@

We instrumented the heterogenous implementation of our interface to 
capture event usage information.  Figure~\ref{fig:eventlife} shows
a timeline of the execution of the Fluid application on 16 nodes.
The dynamic events line measures the total number of event creations.
A large number of events are created - over 260,000 in less than 15
seconds - and allocating separate storage for every event would clearly
be difficult for long-running applications.

Dynamic events are considered {\em live events} until their last operation 
(i.e. wait, trigger) has been performed.  The live events line
measures the number of events that are live at any given time.  After
the last operation on an event, a reference counting implementation would
be able to recover the storage associated with the event.  In this example 
a reference counting algorithm would reduce the storage needed for dynanmic events
by over 10X, but could only do so at the cost of
computation and communication for performing the reference counting.

As discussed in Section~\ref{subsec:eventimpl}, our implementation
requires storage that grows with the maximum number of {\em untriggered events}, a number
that is 10X smaller than even the maximal live event count.  The actual storage
requirements of our implementation are shown by the {\em physical events} line,
which is slightly larger than the peak number of untriggered events. 
The reason for this difference is because nodes create events locally if they have
no free generational events, even if there are unused generational events on remote nodes.
Despite this inefficiency, our implementation uses 5X less storage
than a reference counting implementation and avoids any overhead related to 
reference counting.  These savings would likely be even more dramatic for longer 
runs of the application, as the number of live events is steadily growing as the
application runs, while the peak number of generational events needed appears to 
occur during the startup of the application.

@@

We also instrumented our heterogeneous implementation to profile the usage of 
locks in real applications.  We measured how many locks were used in the application.  For
each lock we measured the number of lock requests and the number of times that the lock
had to be transferred from one node to another.  Figure~\ref{fig:lockcount} shows the results 
for the Circuit and AMR application, both running on 16 nodes.  The Circuit application uses
a larger number of locks (3336 vs. 1393), but most of them (86\%) are only used on the node on 
which they were created.  The AMR application exhibits more varied locking behavior, and in 
particular has locks that are requested more often than they are transferred.  Without the
ability for locks to migrate, these applications couldn't have made use of asynchronous
locks.

@@

Our last set of experiments demonstrates that the asynchronous nature of our interface
is essential to the high performance of the Legion runtime system.  The Legion
implementations of our three applications have already been shown to outperform 
pre-existing implementations\cite{Legion12}.  To quantify the benefit of the asynchronous
nature of our interface, we implemented versions of our applications in
Legion in a bulk-synchronous style to compare against our existing implementations.  
These bulk-synchronous versions still run on top of 
the Legion runtime, but invoke blocking operations between computational phases,
communication phases, and synchronization.

@@

These results show that the asynchronous nature of our interface allows
the Legion runtime system to run ahead of actual execution which
provides significant latency hiding capability.  
The benefit of this latency hiding grows as the number of
nodes increases. % and as the computation-to-communication ratio decreases.
In all cases, the overhead of the bulk-synchronous
implementations grew with the node count.  As we continue to scale
applications to larger and larger machines, the ability of runtimes like Legion to run ahead
will be essential to hiding the large latencies inherent in such machine architectures.
The fully asynchronous nature of the interface presented in this paper
will be crucial for supporting this ability.
